Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method

### 2023-08-31 manu-draft.qmd
update sensitivity analysis results text

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck23]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed. Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at one day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. All results provided herein were generated using the EBASE R package.

@@

Several configurations of the EBASE model can influence how the parameters and metabolic rates are estimated. Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the synthetic time series. First, results were compared to the known values after changing characteristics of their prior distributions. Second, results were compared using a model period of optimization for one, seven, and 30 days to assess how results varied based on the amount of data used for the Bayesian estimation. Details of these analyses are as follows.

@@

The optimization period used for a time series determines how much data are used to estimate the $a$ and $R$ parameters. Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$ and $R$ are returned by the model ($b$ is also returned, but was fixed for the sensitivity analysis). A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period. Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.

Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying by optimization period. @fig-priorcomp shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and the three optimization periods of one, seven, and thirty days. The results are shown using Nash-Sutcliffe Efficiency (NSE) values [@Nash70; @Moriasi07]:

@@

where $Y_i$ is the estimate from EBASE or the synthetic times series for the optimization period $i$, evaluated for each metabolic parameter, each unique combination of priors, and each optimization period. The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for one day, $n \approx 52$ for seven days, and $n \approx 12$ for 30 days. All metabolic estimates were averaged within the number of days for each optimization period prior to calculating NSE values, if appropriate. The NSE value is conceptually similar to the coefficient of determination, but varies from $-\infty$ to 1. Values in the positive range from 0 to 1 are desirable with 1 being a perfect fit, whereas values in the negative range indicate the mean of the synthetic data is a better predictor than EBASE.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $mu$ for one of the $a$ and $R$ parameters was high. However, substantial variation in NSE values was observed among all other parameters. Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`). Conversely, the $a$ parameter was reproduced the most poorly, with a median value less than zero across all optimization periods and priors (median NSE = `r prmall[['a']]`). By optimization period, median NSE values across all parameters increased as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`). The top three ranked models were identical across optimization periods and median NSE values across parameters for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`). The NSE values for all parameters were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the one day optimization period. About half of the parameters regardless of optimization period or priors had NSE \< 0 (`r perneg`). As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$, although the second and third best performing models had high $\sigma$ for $R$.

@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters. Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown. For example, the left-most bars in top left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\mu$ value for the $a$ prior distribution was low or high. The lines indicate the inter-quartile range of NSE values for each subset. As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter. Dissolved oxygen was generally modeled with high accuracy for all priors and it was excluded to not skew the assessment. The patterns suggest that EBASE results are more sensitive to model optimization period as shown by higher median NSE values with longer optimization times (note the different y-axis scales, also confirmed in @fig-priorplot). Similarly, median values were much more variable for the one day optimization period, as shown by a mix of both positive and negative values in the interquartile ranges. Lower median NSE values were observed for higher values for $\mu$, where higher median values for were observed for higher values of $\sigma$, regardless of optimization period.

The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability). The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameter estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period. The worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values). The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplot b, d, and f.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism, as well as alternative statistical approaches to estimating the parameters. Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE were uninformed using the same large $\sigma$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ for $a$, $\mu$ = 20 $\text{mmol}/\text{m}^3/\text{d}$ for $R$, $\mu$ = 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ for $b$). A one day model optimization period was used for EBASE for comparability with the Odum method.

@@

EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23]. Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for a single day. Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters. Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., wind patterns that can affect gas exchange as quantified by the $b$ parameter may persist for several days. Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period. However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location. Our example using the synthetic time series from Apalachicola Bay suggests that results were similar for the 7 and 30 day optimization period. As such, using one week of data for each optimization period is likely appropriate for this location. For other locations, using one day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates and will likely produce parameters that are overly generalized.

@@

![Comparison of EBASE results to a synthetic time series created with known model parameters. ($a$, $R$, and $b$). Results are compared for unique combinations of prior distributions and optimization period. The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is one day, seven days, or thirty days. The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. The $b$ parameter is a fixed constant and was not evaluated. Rankings of each combination as the median NSE value across all five model outputs in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

@@

![Summaries of Nash-Sutcliffe Efficiency (NSE) values across all the comparisons of EBASE and the five synthetic model outputs separated by different combinations of prior values (note the different y-axis scales). Each result on the x-axis represents the median values for which the combination prior applies. For example, the bars in the upper left panel that are farthest to the left summarize the results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period. Bars are the medians and lines the interquartile ranges of NSE values for each result ($n$ = 8 for each bar) .](figs/priorsumcomp.png){#fig-priorsumcomp}

@@

![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters. The best and worst performing models for the one day (a, b), seven day (c, d), and thirty day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficency values across all parameter estimates (@fig-priorcomp). The results are averaged at the time step of the model optimization period.](figs/optex.png){#fig-optex fig-align="center"}

>>>>

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck23]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed. Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. All results provided herein were generated using the EBASE R package.

@@

Several configurations of the EBASE model can influence how the parameters and metabolic rates are estimated. Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the synthetic time series. First, results were compared to the known values after changing characteristics of their prior distributions. Second, results were compared using a model period of optimization for 1, 7, and 30 days to assess how results varied based on the amount of data used for the Bayesian estimation. Details of these analyses are as follows.

@@

The optimization period used for a time series determines how much data are used to estimate the $a$ and $R$ parameters. Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being 1 day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$ and $R$ are returned by the model ($b$ is also returned, but was fixed for the sensitivity analysis). A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period. Results from optimization periods of 1, 7, and 30 days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.

Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying by optimization period. @fig-priorcomp shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and the three optimization periods of 1, 7, and 30 days. The results are shown using Nash-Sutcliffe Efficiency (NSE) values [@Nash70; @Moriasi07]:

@@

where $Y_i$ is the estimate from EBASE or the synthetic times series for the optimization period $i$, evaluated for each metabolic parameter, each unique combination of priors, and each optimization period. The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for 1 day, $n \approx 52$ for 7 days, and $n \approx 12$ for 30 days. All metabolic estimates were averaged within the number of days for each optimization period prior to calculating NSE values, if appropriate. The NSE value is conceptually similar to the coefficient of determination, but varies from $-\infty$ to 1. Values in the positive range from 0 to 1 are desirable with 1 being a perfect fit, whereas values in the negative range indicate the mean of the synthetic data is a better predictor than EBASE.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $mu$ for one of the $a$ and $R$ parameters was high. However, substantial variation in NSE values was observed among all other parameters. Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`). Conversely, the $a$ parameter was reproduced the most poorly, with a median value less than zero across all optimization periods and priors (median NSE = `r prmall[['a']]`). By optimization period, median NSE values across all parameters increased as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`). The top three ranked models were identical across optimization periods and median NSE values across parameters for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`). The NSE values for all parameters were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the 1 day optimization period. About half of the parameters regardless of optimization period or priors had NSE \< 0 (`r perneg`). As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$, although the second and third best performing models had high $\sigma$ for $R$.

@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters. Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown. For example, the left-most bars in top left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a 1-day optimization period where the $\mu$ value for the $a$ prior distribution was low or high. The lines indicate the inter-quartile range of NSE values for each subset. As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter. Dissolved oxygen was generally modeled with high accuracy for all priors and it was excluded to not skew the assessment. The patterns suggest that EBASE results are more sensitive to model optimization period as shown by higher median NSE values with longer optimization times (note the different y-axis scales, also confirmed in @fig-priorplot). Similarly, median values were much more variable for the 1 day optimization period, as shown by a mix of both positive and negative values in the interquartile ranges. Lower median NSE values were observed for higher values for $\mu$, where higher median values for were observed for higher values of $\sigma$, regardless of optimization period.

The actual model results for the best and worst performing combinations of prior distributions for the 1 day, 7 day, and 30 day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability). The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameter estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the 1 day optimization period. The worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values). The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplot b, d, and f.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism, as well as alternative statistical approaches to estimating the parameters. Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE used the default values justified in @sec-ebasemod (equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.

@@

EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23]. Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for a single day. Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters. Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., wind patterns that can affect gas exchange as quantified by the $b$ parameter may persist for several days. Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period. However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location. Our example using the synthetic time series from Apalachicola Bay suggests that results were similar for the 7 and 30 day optimization period. As such, using one week of data for each optimization period is likely appropriate for this location. For other locations, using 1 day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates and will likely produce parameters that are overly generalized.

@@

![Comparison of EBASE results to a synthetic time series created with known model parameters. ($a$, $R$, and $b$). Results are compared for unique combinations of prior distributions and optimization period. The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is 1 day, 7 days, or 30 days. The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. The $b$ parameter is a fixed constant and was not evaluated. Rankings of each combination as the median NSE value across all five model outputs in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

@@

![Summaries of Nash-Sutcliffe Efficiency (NSE) values across all the comparisons of EBASE and the five synthetic model outputs separated by different combinations of prior values (note the different y-axis scales). Each result on the x-axis represents the median values for which the combination prior applies. For example, the bars in the upper left panel that are farthest to the left summarize the results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a 1 day optimization period. Bars are the medians and lines the interquartile ranges of NSE values for each result ($n$ = 8 for each bar) .](figs/priorsumcomp.png){#fig-priorsumcomp}

@@

![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters. The best and worst performing models for the 1 day (a, b), 7 day (c, d), and 30 day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficency values across all parameter estimates (@fig-priorcomp). The results are averaged at the time step of the model optimization period.](figs/optex.png){#fig-optex fig-align="center"}
